---
title: "02. Descriptive statistics"
author: "COADDS24.1F-011/030/032"
output: 
  html_document:
    css: ../reports/styles.css
    code_folding: show
knit: (function(inputFile, encoding) {
      rmarkdown::render(inputFile, output_dir = "../reports")})
---
# Loading libraries
```{r}
source("functions.R")

required_packages <- c("dplyr", "ggplot2", "cowplot", "GGally", "rlang", "corrplot", "RColorBrewer")

for (package in required_packages) {
  if (!suppressPackageStartupMessages(require(package, character.only = TRUE))) {
    install.packages(package)
  }
  suppressPackageStartupMessages(library(package, character.only = TRUE))
  cat("loaded package:", package,"\n")
}
```
# Separating Train and Test data
```{r}
df <- read.csv("../data/01_raw_dataframe.csv", header=TRUE)

set.seed(91)

train <- sample_frac(df, 0.8)

train.data <- as.numeric(rownames(train))
test <- df[-train.data,]

write.csv(train, "../data/02_raw_train_data.csv", row.names = FALSE)
write.csv(test, "../data/03_raw_test_data.csv", row.names = FALSE)
```
an 80% - 20% split was used between training and testing data (using dplyr)

# Checking variables and data types
```{r}
glimpse(train)
```
## Insights:

* the train dataset is still large (2,864 rows) and will likely create an accurate model
* Data is quantitative and continuous
* All data types are double, which is needed for accuracy

# Checking Summary
```{r}
summary(train)
```

## Insights:

* Ranges of data are different, scaling might be needed at some step of the preprocessing.

# Univariate Analysis

```{r, message=FALSE, warning=FALSE}
create_plots(train, info_univariate,"histogram", "01", "histograms")
create_plots(train, info_univariate,"density", "01", "density_plots")
create_plots(train, info_univariate,"boxplot", "01", "boxplots", 4)
create_plots(train, info_univariate,"qq", "01", "Q-Q_plots")
```

## Insights:

* All variables may have outliers.
* They are not similar to normal gaussian distributions.
* Most distributions are skewed.
* Humidity data has the best quality (closest to normal distribution, less outliers, less skewness).
* Light_level data has the worst quality.
* Q-Q plots show that variables do not have a significant similarity to a normal distribution.

# Bivariate Analysis

```{r, message=FALSE, warning=FALSE}
create_bivariate_plot(train, info_bivariate, "01")
```

## Insights:

* Temperature and humidity variables have a high negative correlation of -0.9013
* Temperature and uv_level variables have a low positive correlation of 0.3840
* Temperature and light_level variables have a very low positive correlation of 0.2337
* Once again it is visible that humidity has the best data quality (low residuals and high correlation).
* Expected outcome of performing a multiple linear regression is that light_level will show no statistically significant relationship but the other variables will.

# Multivariate analysis

```{r, message=FALSE, warning=FALSE}
create_multivariate_plot(train, info_univariate, "01")
```

## Insights:

* Since all variables have different units, they take different ranges.
* To make the data usable, scaling will have to be done.


