---
title: "04. Model fitting and checking assumptions"
author: "COADDS24.1F-011/030/032"
output: 
  html_document:
    css: ../reports/styles.css
    code_folding: show
knit: (function(inputFile, encoding) {
      rmarkdown::render(inputFile, output_dir = "../reports")})
---

# Loading libraries

```{r}
source("functions.R")

required_packages <- c("ggplot2", "lmtest", "car")

for (package in required_packages) {
  if (!suppressPackageStartupMessages(require(package, character.only = TRUE))) {
    install.packages(package)
  }
  suppressPackageStartupMessages(library(package, character.only = TRUE))
  cat("loaded package:", package,"\n")
}
```

# Analysis of the fitted model
```{r}
df <- read.csv("../data/06_scaled.csv", header=TRUE)

model <- lm(temperature ~ humidity + light_level + uv_level, data = df)

print(summary(model, diagnostics=TRUE))
```
## Summary Interpretation

* The MLR model is represented as $$\hat{Y} = \beta_0 + \beta_1 X_1 + \beta_2 X_2 + \beta_3 X_3 + \epsilon$$
where,\
$\hat{Y}$ = Predicted Temperature\
$\beta_0$ = y intercept term\
$X_1, X_2, X_3$ = independent variables (Humidity,Light intensity,UV intensity)\
$\beta_1, \beta_2, \beta_3$ =  regression coefficients of Humidity,Light intensity,UV intensity \
$\epsilon$ = error term/residuals

* Humidity has a $p-value < 0.001$ (Signif.code: '\*\*\*' [0, 0.001]). therefore the humidity variable shows a high statistically significant relationship with temperature at the 5% significance level.

* null hypothesis for $X_1$ predictor is rejected at 95% confidence level, accept $𝐻_1: \beta_1 ≠ 0$

* light_level has a $p-value > 0.05$ (Signif.code:' '[0.05, 1]). Therefore, the light_level variable does not show a statistically significant relationship with temperature at the 5% significance level.

* null hypothesis for $X_2$ predictor is accepted at 95% confidence level, accept $𝐻_0: \beta_2 = 0$

* uv_level has a $p-value < 0.001$ (Signif.code: '\*\*\*' [0, 0.001]). therefore the uv_level variable shows a high statistically significant relationship with temperature at the 5% significance level.

* null hypothesis for $X_3$ predictor is rejected at 95% confidence level, accept $𝐻_1: \beta_3 ≠ 0$

* Overall null hypothesis is rejected $(H_0: \beta_0 = \beta_1 = \beta_2 = \beta_3 = 0)$. accept alternate hypothesis

### Conclusion

Humidity and uv_level variables show high statistically significant relationships with temperature while light_level does not show a statistically significant relationship with temperature at the 5% significance level. $$H_1: \exists \beta_i \neq 0, \text{ for at least one } i \in \{1,2,3\}$$.

Therefore the final regression model is interpreted as $$\hat{Y} = (0.724744) + (-0.888851) X_1 + (-0.002041) X_2 + (0.144616) X_3 \pm (0.0611)$$

## Residuals vs Fitted Plot (Linearity)
```{r}
ggplot_data <- data.frame(fitted = model$fitted.values, residuals = model$residuals)
p <- ggplot(ggplot_data, aes(x = fitted, y = residuals)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linewidth = 1) +
  labs(x = "Fitted Values", y = "Residuals", title = "Residuals vs Fitted")

ggsave("../figures/04_residuals_vs_fitted_plot.png", plot = p, width = 6, height = 4, dpi = 300)
print(p)
```

### Objective
to check if there is a linear relationship between predictors and the response (linearity).

### Conclusion
Residuals are scattered but lie close to the 0 residual line (accurate values)\
Therefore, The model captures the trend correctly and errors are randomly distributed. There is a linear relationship between predictors and the response.

## Durbin-Watson Test (Independence)
```{r}
lmtest::dwtest(model)
```

### Objective

to check if the error terms are independent or if they exhibit a pattern over time (autocorrelation)

* $H_0$ - residuals have no autocorrelation
* $H_1$ - true autocorrelation is greater than 0

### Conclusion

$H_0$ is accepted at $\alpha = 0.05$ (95% confidence level) since $p-value > 0.05$, residuals have no autocorrelation.

## QQ Plot and Shapiro-Wilk Test for Residuals (Normality)
```{r}
residuals_df <- data.frame(residuals = model$residuals)
p <- ggplot(residuals_df, aes(sample = residuals)) +
  geom_qq() +
  geom_qq_line(color = "red", linewidth = 1) +
  labs(title = "QQ Plot of Residuals", x = "Theoretical Quantiles", y = "Sample Quantiles")

ggsave("../figures/04_residual_Q-Q_plot.png", plot = p, width = 6, height = 4, dpi = 300)
print(p)

shapiro.test(model$residuals)
```

### Objective

to check whether the distribution of samples follow a normal distribution\

* $H_0$ - the residuals are not significantly different from a normal population
* $H_1$ - the residuals are different from a normal population

### Conclusion

$H_0$ is rejected at $\alpha = 0.05$ (95% confidence level) since $p-value < 0.05$, residuals do not follow a normal distribution.

### Q-Q plot interpretation
Although the shapiro test does not confirm normality, the residual points mostly fall on the Q-Q plot reference line, therefore it can be assumed that residuals follow a normal distribution

## non constant variance test (Homomscedasticity)
```{r}
ncvTest(model)
```

### Objective

to check the if the variance of residuals are equal (checking for homoscedasticity)\

* $H_0$ - Homoscedasticity is present (the residuals have constant variance)
* $H_a$ - Homoscedasticity is not present (the residuals do not have constant variance)

### Conclusion

$H_0$ is accepted at $\alpha = 0.05$ (95% confidence level) since $p-value > 0.05$, residuals have constant variance, MLR model is homoscedastic

## Variance Inflation Factor (VIF) (Multicollinearity)
```{r}
car::vif(model)
```

### Objective

Identifying if multicollinearity exists within independent variables

### Conclusion

for all variables, the $1<VIF<5$ which shows a very low correlation with other predictors which is acceptable . THere is no significant multicollinearity between the predictor variables
